---
title: "Network diffusion-based analysis of omics"
author: "Ettore Mosca"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_width: 7
    fig_heigth: 7
vignette: >
  %\VignetteIndexEntry{Network diffusion-based analysis of omics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Installation
The package can be installed from within R:
```{r, eval=FALSE}
install.packages("/path_to_package/dmfind_xxx.tar.gz", repos = NULL)
library(dmfind)
```

##Input data
The package requires an adjacency matrix and one or more input vectors. Let's create a simulated input dataset, including: a normalized adjacency matrix (W), a 1-column matrix (X0) with a vector representing omics-derived statistics:

```{r, eval=TRUE, message=FALSE}
library(igraph)
library(dmfind)
data(g400esm, W, X0, package="dmfind")

summary(g400esm)
head(X0)

```
Now, we need to explore a bit the possible values of the $\epsilon$ value. Let's apply network diffusion to our input vector:
```{r, eval=TRUE, message=FALSE}
library(dmfind)
Xs <- ND(X0, W)$Xs
rank_X0 <- rank(-X0, ties.method = "min")
rank_Xs <- rank(-Xs, ties.method = "min")
par(mar=c(4, 4, 2, 1))
par(mfrow=c(1, 2))
L <- layout_nicely(g400esm)
plot(X0[, 1], Xs[, 1], pch=16)
text(X0[, 1], Xs[, 1], V(g400esm)$label[match(rownames(X0), V(g400esm)$name)], pos = 1, cex=0.8)
V(g400esm)$shape <- ifelse(rank(-X0[match(V(g400esm)$name, rownames(Xs)), 1])<=6, "square", "circle")
V(g400esm)$color <- heat.colors(nrow(Xs))[rank_Xs[match(V(g400esm)$name, rownames(X0))]]
par(mar=c(0.1, 0.1, 0.1, 0.1))
plot(g400esm, vertex.size=8, vertex.label.cex=0.5,
     edge.width=1.5, edge.color="hotpink",
     vertex.label=rank_Xs[match(V(g400esm)$name, rownames(X0))],
     vertex.label.cex=1.3,
     layout=L
)

```
As expected, network based scores prioritize genes in network proximity to high $X_0$ values. The initial and final state of the system are comapred by means of the network smoothing index $S$, which require a tuning parameter $\epsilon$. Briefly, low $\epsilon$ highliths genes with the highest gain, while for $\epsilon \to \inf$ the same gene ranking is obtained by $S$ and $X_s$. We ca explore the realtion between $\epsilon$ and $S$ with `eval_eps` function.
```{r, eval=TRUE, message=FALSE}
eps <- matrix(c(0.01, 0.1, 1), ncol=1)
out <- eval_eps(X0, Xs, eps, g400esm, top=20)

par(mfrow=c(1, 2))
eps_text <- format(eps, digits = 2)
for(i in 1:nrow(eps)){
  rank_Si <- rank(-out$S[[i]], ties.method = "min")
  par(mar=c(4, 4, 3, 3))
  plot(X0[, 1], out$S[[i]][, 1], xlab="X0", ylab="S", main=eps_text[i], pch=16)
  text(X0[, 1], out$S[[i]][, 1], V(g400esm)$label[match(rownames(X0), V(g400esm)$name)], pos = 1, cex=0.8)
  V(g400esm)$color <- heat.colors(nrow(out$S[[i]]))[rank_Si[match(V(g400esm)$name, rownames(out$S[[i]]))]]
  par(mar=c(0.1, 0.1, 0.1, 0.1))
  plot(g400esm, vertex.size=8, vertex.label.cex=0.5,
       edge.width=1.5, edge.color="hotpink",
       layout=L
  )
}
```
It's useful to check the $\Omega$ function calcualed on $X_0$ and $X_s$ values using the ranking determined by $S(\epsilon)$.
```{r, eval=TRUE, message=FALSE}
par(mfrow=c(1, 2))
par(mar=c(4, 4, 3, 3))
plot(out$csX0[[1]][, 1], type='l', xlab="rank", ylab="omega(X0) ranked by S", ylim=c(0, max(unlist(lapply(out$csX0, function(x) x[, 1])))), col=1, lty=2)
for(i in 2:nrow(eps)){
  lines(out$csX0[[i]][, 1], col=i, lty=2)
}
legend("topleft", legend = format(eps[, 1], digits = 2), lty=1, col=1:nrow(eps), xpd=NA, cex=0.5)
plot(out$csXs[[1]][, 1], type='l', xlab="rank", ylab="omega(Xs) ranked by S", ylim=c(0, max(unlist(lapply(out$csXs, function(x) x[, 1])))), col=1, lty=2)
for(i in 2:nrow(eps)){
  lines(out$csXs[[i]][, 1], col=i, lty=2)
}
legend("topleft", legend = format(eps[, 1], digits = 2), lty=1, col=1:nrow(eps), xpd=NA, cex=0.5)

```
We can see that, considering the given initial conditions $(X_0, W)$, $\epsilon=0.01$ determines a gene ranking with lower $\Omega$ both in relation at $X_0$ and $X_s$, while higher $\Omega$ can be obtained using higher $\epsilon$ values. For this tutorial we will therefore use $\epsilon=0.1$.

We can assess the statistical significance of $S$ using empirical $p$-values.
```{r, eval=TRUE, message=FALSE}
out <- calc_adjND(X0, W, eps = 0.1, k = 99)
rank_Sp <- rank(-out$Sp)
par(mfrow=c(1, 2))
par(mar=c(4, 4, 1, 1))
#plot(out$S, -log10(out$p), xlab="S", ylab="-log10(p)", pch=16)
plot(X0, out$Sp, xlab="X0", ylab="Sp", pch=16)
text(X0, out$Sp, V(g400esm)$label[match(rownames(X0), V(g400esm)$name)], pos = 1, cex=0.8)

V(g400esm)$color <- heat.colors(nrow(out$Sp))[rank_Sp[match(V(g400esm)$name, rownames(out$Sp))]]
par(mar=c(0.1, 0.1, 0.1, 0.1))
plot(g400esm, vertex.size=8, vertex.label.cex=0.5,
     edge.width=1.5, edge.color="hotpink",
     layout=L
)
```

Network resampling can be used to asses the presence of significantly connected modules. 
```{r, eval=TRUE, message=FALSE}
out_nr <- NR(g400esm, ranked_vector = as.array(sort(out$Sp[, 1], decreasing = T))[1:30], k = 99, null_model = "A")
par(mfrow=c(1, 2))
par(mar=c(4, 4, 1, 1))
plot(out_nr$omega_perm[, 1], xlab="rank", ylab="omega", type="l", col="gray", ylim = c(0, max(c(out_nr$NR_summary$omega, unlist(out_nr$omega_perm)))), lty=2, lwd=2)
for(i in sample(2:99, 40)){
  lines(out_nr$omega_perm[, i], col="gray", lty=2)  
}
lines(out_nr$NR_summary$omega, col="red")
plot(log10(out_nr$NR_summary$p), xlab="rank", ylab="log10(p)", type="b", pch=16)

out_nr <- NR(g400esm, ranked_vector = as.array(sort(out$Sp[, 1], decreasing = T))[1:30], k = 99, null_model = "S")
par(mfrow=c(1, 2))
par(mar=c(4, 4, 1, 1))
plot(out_nr$omega_perm[, 1], xlab="rank", ylab="omega", type="l", col="gray", ylim = c(0, max(c(out_nr$NR_summary$omega, unlist(out_nr$omega_perm)))), lty=2, lwd=2)
for(i in sample(2:99, 40)){
  lines(out_nr$omega_perm[, i], col="gray", lty=2)  
}
lines(out_nr$NR_summary$omega, col="red")
plot(log10(out_nr$NR_summary$p), xlab="rank", ylab="log10(p)", type="b", pch=16)

out_nr <- NR(g400esm, ranked_vector = as.array(sort(out$Sp[, 1], decreasing = T))[1:30], k = 99, null_model = "SA")
par(mfrow=c(1, 2))
par(mar=c(4, 4, 1, 1))
plot(out_nr$omega_perm[, 1], xlab="rank", ylab="omega", type="l", col="gray", ylim = c(0, max(c(out_nr$NR_summary$omega, unlist(out_nr$omega_perm)))), lty=2, lwd=2)
for(i in sample(2:99, 40)){
  lines(out_nr$omega_perm[, i], col="gray", lty=2)  
}
lines(out_nr$NR_summary$omega, col="red")
plot(log10(out_nr$NR_summary$p), xlab="rank", ylab="log10(p)", type="b", pch=16)

```

The variation of the empirical $p$-value with gene rank suggests modules with high scores in network proximity. Note that this procedure is designed to focus on the top of a gene list, in this small example the top 30 genes out of a network of 61.
